Black-hole ringdown as a probe of higher-curvature gravity theories

### 2022-04-22 papers/paper_alt_theor_bounds/paper_alt_theor_bounds.tex
adding contour plot discussion to dcs section

\begin{figure}[t]
\includegraphics[width=0.9\columnwidth]{figs/tmp_GW150914_intrinsic_params_remnant.pdf}
\caption{(Color online).~Corner plot showing that the inferred final spin $\chi_f$,
remnant mass $M_{\rm f}$ for GW150914, using the same waveform model,
but without (purple contours) and with the non-GR parameters
different from zero (blue contours), for
dCS gravity and $N_{\rm max}=0$ as an example.
%
The contours represent 90\% credible levels.
%
We see that the introduction of the non-GR parameters does not bias the
inference on the source parameters.
}
\label{fig:corner_plot}
\end{figure}

@@

\begin{table}[h]
\begin{tabular}{l l c c c}
\hline
\hline
$N_{\rm max}$ & Event &  EFT    & ParSpec & Constraint      \\
              &       &  bound? & bound?  & ($\gm = M_{f}$) \\
\hline
  & GW150914 & No  & Yes & -- \\
0 & GW200129 & No  & Yes & -- \\
  & Combined & --  & Yes & -- \\
\hline
  & GW150914 & No  & Yes & -- \\
1 & GW200129 & No  & Yes & -- \\
  & Combined & --  & Yes & -- \\
\hline
\hline
\end{tabular}
\caption{Detailed summary of our results for EdGB gravity for GW150914, GW200129, and
combined events using $\gm = M_{\rm f}$, $\varepsilon = 1/2$ and quoting only 90\% credible results.
%
We find that we cannot place any constraint on $\ell_{\rm EdGB}$ with our waveform model from either GW event.
}
\label{tab:summary_edgb}
\end{table}

@@

In Fig.~\ref{fig:dcs_cdf} we show the CDF for GW150914, we see that with $\gm = m_2$,
Eq.~\eqref{eq:eft_bound} is not satisfied unless $\varepsilon \approx 0.9$ (with only $j=0$ corrections) and
$\varepsilon \approx 0.7$ (with both $j=0$ and $1$ corrections).
%
The situation is different if we use $\gm = M_{\rm f}$. In this case, we find that
with or without spin corrections Eq.~\eqref{eq:eft_bound} can be satisfied
with $\varepsilon \leqslant 1/2$ (i.e., below the criteria used
Refs.~\cite{Nair:2019iur,Perkins:2021mhb,Lyu:2022gdr}).
%
This means that with our model's assumptions and using the remnant's source mass $M_{\rm f}$ to set the
cut-off scale that we can claim an upper bound
%
\begin{equation}
\ell_{\rm dCS} \leqslant 42.6~\textrm{km}
\quad \textrm{at 90\% credible level},
\end{equation}
%
on dCS gravity and would constitute \emph{the first bound on this theory with GW
observations alone}.
%
We also note that since the CDF grows fast in approximate domain $\varepsilon = [0.2, 0.4]$,
that a stronger, albeit at lower credibility,
%
\begin{equation}
\ell_{\rm dCS} \leqslant 37.4~\textrm{km}
% ~( = \Lambda_{\rm EFT}(0.3, M_{f}) )
\quad \textrm{at 68\% credible level},
\end{equation}
%
can also be placed, when $N_{\rm max}=0$.

@@

@@

\sout{In Fig.~\ref{fig:corner_plot}, we show the final spin $\chi_f$ and
detector-frame remnant mass $(1 + z) M_{\rm f}$ for GW150914 for GR and the
illustrative case of dCS gravity. We see that our waveform models does not
introduce a substantial change to the GR estimates on this parameters, as
required by the ParSpec expansion (see discussion in
Sec.~\ref{sec:review_parspec}).}\agcomm{This reference to Fig.2 appears misplaced here. IMO, Fig 2. has two purposes. 1) to show that we do not need the high-spin version of the fits Carullo used, because our spin posteriors are well-localised around 0.7, and 2) to assure the reader that we were right in using the GR values of the final mass and spin in the ParSpec expansion, because introducing nonGR parameters has not ended up biasing our estimates of Mf-chif.  I think both of these points can be addressed in the discussions section. That is also where we can put the other corner plot with the remaining intrinsic parameters. One side-advantage of this is that the figures would then also appear slightly later in the paper, as we had originally planned.}
%

@@

We presented an unified framework that combines the ParSpec formalism introduced
to model deviations to the GR QNMs~\cite{Maselli:2019mjd} with the \pSEOB~waveform model~\cite{Brito:2018rfr,Ghosh:2021mrv}.
%
We showed with concrete examples, how theory-specific QNM calculations of
slowly-rotating BHs in modified gravity theories can be mapped onto the
non-GR parameters of the ParSpec formalism.
%
Put together this allowed us to use the \pSEOB~waveform model to test four
modified gravity theories (EdGB, dCS, cubic, and quartic EFTs of GR) using
observational data from the LVK events GW150914 and GW200129.
%
We found, in particular that, within the interpretation of our non-GR theories as EFTs and the
region of validity of the ParSpec framework, the coupling constant of dCS gravity is bound to
$\ell_{\rm dCS} \leqslant 34.5$~km, at 90\% credible level, when stacking the posteriors
of GW150914 and GW200129.
%
In contrast, we could not place any constraint of the coupling constants in EdGB gravity $\ell_{\rm EdGB}$.
%
This dichotomy between these theories has an analogy with works that considered
the inspiral part of the GW signal alone.
%
In that case, it is dCS which is \ab{poorly constrained with current observed GW
events}, while EdGB can be constrained, as shown in Refs.~\cite{Nair:2019iur,Perkins:2021mhb,Lyu:2022gdr}.
%
\hscomm{I will add a sentence here on Refs.~\cite{Alexander:2017jmt,Loutrel:2018ydv,Loutrel:2018rxs}.}

@@

We now expand Eq.~\eqref{eq:omega_dsc_eg} to leading-orders in $\chi$ and $\ell_{\rm dCS}$, and
gather the terms proportional to $\ell_{\rm dCS}$.
%
We obtain
%
\begin{align}
    M \, \omega_{\rm dCS} &=
    \left[ 0.3722 + 1.1945 (\ell_{\rm dCS} / M)^4 \right]
    \nonumber \\
    &\quad + \left[0.1861 + 5.1828 (\ell_{\rm dCS} / M)^4 \right] \, \chi\,,
    \label{eq:omega_dcs_interm}
\end{align}
%
where we made use of the numerical value of the coefficients $c_{i}$.
%
We find that (reassuringly) that the nonrotating GR part of the expression above agree with $\omega^{(0)}$ of Ref.~\cite{Maselli:2019mjd}
to 0.5\% relative percent error. The same error estimate is larger ($\approx 20$\%) for the linear-in-spin coefficient;
0.1861 in comparison to 0.1258 of Ref.~\cite{Maselli:2019mjd}.
%
We attribute this difference to Ref.~\cite{Wagle:2021tam} having fitted Eq.~\eqref{eq:omega_dsc_eg} to QNM data compute to linear-order in spin, whereas~\cite{Maselli:2019mjd}
fitted Eq.~\eqref{eq:kerr_expansion} to Kerr QNM valid to all orders in spin.

>>>>

@@

\begin{table}[b]
\begin{tabular}{l l c c c}
\hline
\hline
$N_{\rm max}$ & Event &  EFT    & ParSpec & Constraint      \\
              &       &  bound? & bound?  & ($\gm = M_{f}$) \\
\hline
  & GW150914 & No  & Yes & -- \\
0 & GW200129 & No  & Yes & -- \\
  & Combined & --  & Yes & -- \\
\hline
  & GW150914 & No  & Yes & -- \\
1 & GW200129 & No  & Yes & -- \\
  & Combined & --  & Yes & -- \\
\hline
\hline
\end{tabular}
\caption{Detailed summary of our results for EdGB gravity for GW150914, GW200129, and
combined events using $\gm = M_{\rm f}$, $\varepsilon = 1/2$ and quoting only 90\% credible results.
%
We find that we cannot place any constraint on $\ell_{\rm EdGB}$ with our waveform model from either GW event.
}
\label{tab:summary_edgb}
\end{table}

@@

In Fig.~\ref{fig:dcs_cdf} we show the CDF for GW150914, we see that with $\gm = m_2$,
Eq.~\eqref{eq:eft_bound} is not satisfied unless $\varepsilon \approx 0.9$ (with only $j=0$ corrections) and
$\varepsilon \approx 0.7$ (with both $j=0$ and $1$ corrections).
%
The situation is different if we use $\gm = M_{\rm f}$. In this case, we find that
with or without spin corrections Eq.~\eqref{eq:eft_bound} can be satisfied
with $\varepsilon \leqslant 1/2$ (i.e., below the criteria used
Refs.~\cite{Nair:2019iur,Perkins:2021mhb,Lyu:2022gdr}).
%
This means that with our model's assumptions and using the remnant's source mass $M_{\rm f}$ to set the
cut-off scale that we can claim an upper bound
%
\begin{equation}
\ell_{\rm dCS} \leqslant 42.6~\textrm{km}
\quad \textrm{at 90\% credible level},
\end{equation}
%
on dCS gravity and would constitute \emph{the first bound on this theory with GW
observations alone}.
%
We also note that since the CDF grows fast in approximate domain $\varepsilon = [0.2, 0.4]$,
that a stronger, albeit at lower credible level,
%
\begin{equation}
\ell_{\rm dCS} \leqslant 37.4~\textrm{km}
\quad \textrm{at 68\% credible level},
\end{equation}
%
can also be placed, when $N_{\rm max}=0$.

@@

\begin{figure}[t]
\includegraphics[width=0.9\columnwidth]{figs/tmp_GW150914_intrinsic_params_remnant.pdf}
\caption{(Color online).~Corner plot showing that the inferred final spin $\chi_f$,
remnant mass $M_{\rm f}$ for GW150914, using the same waveform model,
but without (purple contours) and with the non-GR parameters
different from zero (blue contours), for
dCS gravity and $N_{\rm max}=0$.
%
The contours represent 90\% credible levels.
%
We see that the introduction of the non-GR parameters does not bias the
inference on the source parameters as required by the ParSpec.
}
\label{fig:corner_plot}
\end{figure}

@@

\hscomm{Alessandra, I moved the plot here as we discussed.}
%
In Fig.~\ref{fig:corner_plot}, we show the final spin $\chi_{\rm f}$ and
remnant mass $M_{\rm f}$ for GW150914 for GR and dCS gravity.
%
We see that indeed our waveform models does not introduce substantial changes
to the GR estimates on these parameters, as required by the ParSpec expansion
(see discussion in Sec.~\ref{sec:review_parspec}).
%
We observed the same in all other theories we studied.

% \agcomm{This reference to Fig.2 appears misplaced here. IMO, Fig 2. has two purposes. 1) to show that we do not need the high-spin version of the fits Carullo used, because our spin posteriors are well-localised around 0.7, and 2) to assure the reader that we were right in using the GR values of the final mass and spin in the ParSpec expansion, because introducing nonGR parameters has not ended up biasing our estimates of Mf-chif.  I think both of these points can be addressed in the discussions section. That is also where we can put the other corner plot with the remaining intrinsic parameters. One side-advantage of this is that the figures would then also appear slightly later in the paper, as we had originally planned.}
%

@@

We presented an unified framework that combines the ParSpec formalism introduced
to model deviations to the GR QNMs~\cite{Maselli:2019mjd} with the \pSEOB~waveform model~\cite{Brito:2018rfr,Ghosh:2021mrv}.
%
We showed with concrete examples, how theory-specific QNM calculations of
slowly-rotating BHs in modified gravity theories can be mapped onto the
non-GR parameters of the ParSpec formalism.
%
Put together this allowed us to use the \pSEOB~waveform model to test four
modified gravity theories (EdGB, dCS, cubic, and quartic EFTs of GR) using
observational data from the LVK events GW150914 and GW200129.
%
We found, in particular that, within the interpretation of our non-GR theories as EFTs and the
region of validity of the ParSpec framework, the coupling constant of dCS gravity is bound to
$\ell_{\rm dCS} \leqslant 34.5$~km, at 90\% credible level, when stacking the posteriors
of GW150914 and GW200129.
%
In contrast, we could not place any constraint of the coupling constants in EdGB gravity $\ell_{\rm EdGB}$.
%
This dichotomy between these theories has an analogy with works that considered
the inspiral part of the GW signal alone.
%
In that case, it is dCS gravity which cannot be constrained with current GW
events, while EdGB can, as shown in Refs.~\cite{Nair:2019iur,Perkins:2021mhb,Lyu:2022gdr}.
%
\hscomm{I will add a sentence here on Refs.~\cite{Alexander:2017jmt,Loutrel:2018ydv,Loutrel:2018rxs}.}

@@

We now expand Eq.~\eqref{eq:omega_dsc_eg} to leading-orders in $\chi$ and $\gamma_{\rm dCS}$, and
gather the terms proportional to $\gamma_{\rm dCS}$.
%
We obtain
%
\begin{align}
    M \, \omega_{\rm dCS} =
    \left( 0.3722 + 1.1945 \gamma_{\rm dCS}^4 \right)
    + \left(0.1861 + 5.1828 \gamma_{\rm dCS}^4 \right) \, \chi,
    \nonumber \\
    \label{eq:omega_dcs_interm}
\end{align}
%
where we made use of the numerical value of the coefficients $c_{i}$.
%
We find that (reassuringly) that the nonrotating GR part of the expression above agree with $\omega^{(0)}$ of Ref.~\cite{Maselli:2019mjd}
to 0.5\% relative percent error. The same error estimate is larger ($\approx 20$\%) for the linear-in-spin coefficient;
0.1861 in comparison to 0.1258 of Ref.~\cite{Maselli:2019mjd}.
%
We attribute this difference to Ref.~\cite{Wagle:2021tam} having fitted Eq.~\eqref{eq:omega_dsc_eg} to QNM data compute to linear-order in spin, whereas~\cite{Maselli:2019mjd}
fitted Eq.~\eqref{eq:kerr_expansion} to Kerr QNM valid to all orders in spin.
